### 2011-04-07 Spatial_adaptation/standing-variation.tex
added integral appendix

Now let $f(t) = vt$, so $h(t) = \frac{ \omega_d v^d }{ d+1 } t^{d+1}$.
The integrals that appear above are of the following form:
\begin{align}
  \int_0^\infty t^c \exp \left( - \alpha t^a - \beta t^b \right) dt 
            &= \left( b^{-1} \beta^{ (1-c)/b } \right) \int_0^\infty u^{(c+1-b)/b} \exp\left( - \alpha \beta^{-a/b} u^{a/b} - u \right) du ,
\end{align}
so it suffices to learn something about the function
\begin{equation}
    G(a,b,x) := \int_0^\infty  t^a \exp\left( -x t^b - t \right) dt .
\end{equation}
We will only be interested in this as a function of $x$; 
the coefficients $a$ and $b$ will merely depend on the dimension and what quantity we are calculating.
We can construct a power series for $G$ in $x$:
\begin{align}
    \partial_x^n G(a,b,x) &= \int_0^\infty (-1)^n t^{a+nb} \exp\left( -x t^b - t \right) dt \\
            &= (-1)^n G(a+nb,b,x) , \quad \mbox{so} \\
    \partial_x^n G(a,b,x) \vert_{x=0} &= \Gamma(a+nb+1) .
\end{align}
So, if everything works out, we will have that
\begin{align}
    G(a,b,x) = \sum_{n \ge 0} \frac{(-x)^n}{n!} \Gamma(a+nb+1) .
\end{align}
Something simliar is noted at the Wikipedia page on the Gaussian integral, and claims such integrals show up in quantum field theory.

>>>>

Now let $f(t) = vt$, so $h(t) = \frac{ \omega_d v^d }{ d+1 } t^{d+1}$.
The integrals that appear above are of the following form:
\begin{align}
  \int_0^\infty t^c \exp \left( - \alpha t^a - \beta t^b \right) dt 
            &= \left( b^{-1} \beta^{ (1-c)/b } \right) \int_0^\infty u^{(c+1-b)/b} \exp\left( - \alpha \beta^{-a/b} u^{a/b} - u \right) du ,
\end{align}
so it suffices to learn something about the function
\begin{equation}
    G(a,b,x) := \int_0^\infty  t^a \exp\left( -x t^b - t \right) dt .
\end{equation}
We will only be interested in this as a function of $x$; 
the coefficients $a$ and $b$ will merely depend on the dimension and what quantity we are calculating.
We can construct a power series for $G$ in $x$:
\begin{align}
    \partial_x^n G(a,b,x) &= \int_0^\infty (-1)^n t^{a+nb} \exp\left( -x t^b - t \right) dt \\
            &= (-1)^n G(a+nb,b,x) , \quad \mbox{so} \\
    \partial_x^n G(a,b,x) \vert_{x=0} &= \Gamma(a+nb+1) .
\end{align}
So, if everything works out, we will have that
\begin{align}
    G(a,b,x) = \sum_{n \ge 0} \frac{(-x)^n}{n!} \Gamma(a+nb+1) .
\end{align}
To check for convergence, note that by Stirling's formula $\Gamma(z) \simeq \sqrt{2\pi/z} (z/e)^z$,
\begin{align}
    \left( \frac{x^n \Gamma(a+nb+1) }{ n! } \right)^{1/n} &\simeq x \frac{ (a+nb+1)^{b+(a+1)/n} }{ (n+1)^{1+1/n} } e^{(1-b)-a/n} \\
        &\simeq C n^{b-1} b^b \\
        &\to 0  \quad \mbox{as } n \to \infty ,
\end{align}
so that the sum converges for all $x$ if $b<1$ (which is the case for us as $b=d/(d+1)$).
Something simliar is noted at the Wikipedia page on the Gaussian integral, and claims such integrals show up in quantum field theory.

### 2011-04-07 standing-patches-paper.tex
added integral appendix

The quantity that is perhaps easiest to compute is the mean time until adaptation.
Fix some geographic location, and let $\tau\ge0$ be the time at which the point is reached by the advantageous mutations.
Then, as shown in Figure XX,
$\tau > t$ if and only if the cone with point at $(x,t)$ and slope $v$ extending back to $t=0$ is empty of successful mutations.
Since we assume these are a Poisson process, 
\[
    \P\{ \tau > t \} = \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 \right) ,
\]
and so we have
\begin{align}
    \E[\tau] % &= \int_0^\infty \P\{ \tau > t \} dt \\
        &= \int_0^\infty \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 \right) dt .
\end{align}

@@

We can also find the mean proportion of space covered by standing variants.
If by time $t$ a geographic location has not yet been reached by the mutation,
the probability it will be reached by time $t+dt$ 
by a standing variant is approximately $2 \lambda_0 \pi vt dt$, 
and the probability it is reached by a new variant is $\lambda \pi v^2 t^2 dt$.
Therefore, as is standard for competing exponentials,
the probability a given location is reached first by a standing variant,
and therefore the mean proportion of space covered  by standing variants,
is
\begin{equation}
    z_0 = \int_0^\infty {2 \lambda_0 \pi v t} \exp \left( - 2 \lambda_0 \pi v t - \lambda \pi v^2 t^2 \right) dt .
\end{equation}

Furthermore, if we define $a_0$ to be the mean area occupied by a typical standing variant, 
then $a_0$ is given by the proportion of the range occupied by standing variants over the mean density of unique standing variants,
i.e.\ $a_0 = z_0 / \lambda_0$.
We can compare this to the corresponding quantity $a_+$ for a new variant, since the formula
$a_0 / a_+ = z_0 / (1-z_0)$.

@@

>>>>

The quantity that is perhaps easiest to compute is the mean time until adaptation.
Fix some geographic location, and let $\tau\ge0$ be the time at which the point is reached by the advantageous mutations.
Then, as shown in Figure XX,
$\tau > t$ if and only if the cone with point at $(x,t)$ and slope $v$ extending back to $t=0$ is empty of successful mutations.
Since we assume these are a Poisson process, 
\[
    \P\{ \tau > t \} = \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 \right) ,
\]
and so we have
\begin{align}
    \E[\tau] % &= \int_0^\infty \P\{ \tau > t \} dt \\
        &= \int_0^\infty \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 \right) dt.
\end{align}
For means of evaluting this integral, see Appendix \ref{apx:integrals}.

@@

We can also find the mean proportion of space covered by standing variants.
If by time $t$ a geographic location has not yet been reached by the mutation,
the probability it will be reached by time $t+dt$ 
by a standing variant is approximately $2 \lambda_0 \pi vt dt$, 
and the probability it is reached by a new variant is $\lambda \pi v^2 t^2 dt$.
Therefore, as is standard for competing exponentials,
the probability a given location is reached first by a standing variant,
and therefore the mean proportion of space covered  by standing variants,
is
\begin{equation}
    z_0 = \int_0^\infty {2 \lambda_0 \pi v t} \exp \left( - 2 \lambda_0 \pi v t - \lambda \pi v^2 t^2 \right) dt .
\end{equation}
To evaluate this integral, again see Appendix \ref{apx:integrals}.

Furthermore, if we define $a_0$ to be the mean area occupied by a typical standing variant, 
then $a_0$ is given by the proportion of the range occupied by standing variants over the mean density of unique standing variants,
i.e.\ $a_0 = z_0 / \lambda_0$.
We can compare this to the corresponding quantity $a_+$ for a new variant through the formula
$a_0 / a_+ = z_0 / (1-z_0)$.

@@

\appendix

\section{Integrals appearing in the text}
    \label{apx:integrals}

In the text at several points appear integrals of the form
\begin{align}
  \int_0^\infty t^c \exp \left( - \alpha t^d - \beta t^{d+1} \right) dt 
\end{align}
where $c$ is a positive integer, $d$ is the dimension, and $\alpha$ and $\beta$ are positive real numbers.
This could be evaluated through standard numerical methods; below we describe a power series expansion.
Changing variables to $u = \beta t^{d+1}$, this becomes
\begin{align}
    \left( {d+1}^{-1} \beta^{ (1-c)/(d+1) } \right) \int_0^\infty u^{(c+1-d)/(d+1)} \exp\left( - \alpha \beta^{-d/(d+1)} u^{d/(d+1)} - u \right) du ,
\end{align}
so it suffices to evaluate the function
\begin{equation}
    G(a,b,x) := \int_0^\infty  t^a \exp\left( -x t^b - t \right) dt ,
\end{equation}
in the case that $a=(c-d+1)/(d+1)$, $b=d/(d+1)$, and $x$ is a function of the demographic paramters.
Since $a$ and $b$ only depend on the dimension and the quantity being computed,
we are interested in $G$ as a function of $x$.
At least in the case $b=d/(d+1)$ it is possible to express $G$ as a finite sum of gamma functions,
but we proceed with a simpler method.
Note that $\partial_x G(a,b,x) = -G(a+b,b,x)$,
and that at $x=0$, the function $G$ is the gamma function $G(a,b,0) = \Gamma(a+1)$.
Therefore, a Taylor series for $G$ would be
\[
    G(a,b,x) = \sum_{n \ge 0} = \sum_{n \ge 0} \frac{(-x)^n}{n!} \Gamma(a+nb+1) .
\]
It is easy to check using Stirling's formula that $\limsup_{n \to \infty} ( x^n \Gamma(a+nb+1)/n! )^{1/n} = 0$
if $b<1$, so the sum converges.
